Maximum Entropy Prediction Markets

### 2014-02-10 WriteUp/maxentpredmkts.tex
some more minor edits
added liquidity adjustment to exp util section

\noindent
Recall that for the generalized LMSR, the cost function $C$ is exactly the log partition function $T$. We are therefore lead to the following clear understanding of a risk-averse agent's behavior in such a market.
%
\begin{theorem}
Suppose an agent has exponential utility with coefficient $a$ and exponential family beliefs with natural parameter $\htheta$. In the generalized LMSR market with current market state $\theta$, the agent's optimal trade $\delta$ moves the state vector to
%
\begin{equation} \label{exp-update}
\theta + \delta = \frac{1}{1+a} \htheta + \frac{a}{1+a} \theta.
\end{equation}
%
\end{theorem}
%
%
\begin{proof}
The agent's optimal trade maximizes its expected utility, or equivalently the certainty equivalent. From Lemma~\ref{lem:exp-util} and the fact that $T = C$, the agent therefore maximizes
%
$$
\log a - T(\htheta - a\delta) + T(\htheta) - aT(\theta+\delta) + aT(\theta). 
$$
This objective is strictly concave, from the strict convexity of $T$. The optimum is therefore characterized by the first-order conditions:
$$
\grad T(\htheta - a\delta) = \grad T(\theta+\delta).
$$
As the gradient map $\grad T$ is one-to-one, this is solved by equating the arguments, which leads to $\delta = (\htheta - \theta) / (1+a)$ and~(\ref{exp-update}).
% 
\end{proof}
%

@@

@@

We will show that if we are able to impose finite initial budgets on the traders and control the market prices based on these budgets, then  it possible to set up the market so that it is prohibitive for damaging traders to participate in the market. Further, the informative traders can be shown to have expected growth in budget so that they are eventually able to move the market prices without restriction.

>>>>

\noindent
Recall that for the generalized LMSR, the cost function $C$ is exactly the log partition function $T$. We are therefore lead to the following clear understanding of a risk-averse agent's behavior in such a market.
%
\begin{theorem} \label{thm:exp-util}
Suppose an agent has exponential utility with coefficient $a$ and exponential family beliefs with natural parameter $\htheta$. In the generalized LMSR market with current market state $\theta$, the agent's optimal trade $\delta$ moves the state vector to
%
\begin{equation} \label{exp-update}
\theta + \delta = \frac{1}{1+a} \htheta + \frac{a}{1+a} \theta.
\end{equation}
%
\end{theorem}
%
%
\begin{proof}
The agent's optimal trade maximizes its expected utility, or equivalently the certainty equivalent. From Lemma~\ref{lem:exp-util} and the fact that $T = C$, the agent therefore maximizes
%
$$
\log a - T(\htheta - a\delta) + T(\htheta) - aT(\theta+\delta) + aT(\theta). 
$$
This objective is strictly concave, from the strict convexity of $T$. The optimum is therefore characterized by the first-order conditions:
$$
\grad T(\htheta - a\delta) = \grad T(\theta+\delta).
$$
As the gradient map $\grad T$ is one-to-one, this is solved by equating the arguments, which leads to $\delta = (\htheta - \theta) / (1+a)$ and~(\ref{exp-update}).
% 
\end{proof}
%

@@

\subsubsection*{Liquidity Adjustment}

In practice the centralized market maker allows itself some control over the \emph{liquidity} in the market, which captures how responsive prices are to trades. To adjust liquidity we consider the parametrized cost $C_{\lambda}(\theta)=\frac{1}{\lambda} C\left(\lambda\theta\right)$. Here $\lambda$ is construed as the \emph{inverse liquidity}, or price responsiveness. A larger setting of $\lambda$ means fewer shares need to be bought to reach the same prices.\footnote{Observe that the liquidity adjustment to the cost function takes the same form as the risk-aversion adjustment to the exponential utility in~(\ref{exp-util}). In convex analysis, this transformation is known as the perspective function~\citep{}.}

%(Discussion, you can skip this paragraph.) This is because for higher $\lambda$ we are making the cost function ``more convex''. Specifically, for $\lambda_1 > \lambda_2$ it's the case that $C_{\lambda_1}$ is an increasing convex transformation of $C(\lambda_2)$. This holds because the ``coefficient of risk aversion'' $A(\theta) = C''(\theta)/C'(\theta)$ increases with $\lambda$ at every point. I've written it assuming single-dimensional $\theta$ but there are more complicated looking generalizations for vector $\theta$. In fact, this coefficient might be the ``right'' notion of liquidity at a point, not the second derivative or largest eigenvalue of the Hessian. We should discuss.

\newcommand{\ttheta}{\tilde{\theta}}
In the context of the generalized LMSR we write $T$ rather than $C$ where $T$ is the log partition function, with liquidity-adjusted version $T_{\lambda}$. Let $\hmu$ be the agent's mean belief with corresponding natural parameter $\htheta = \grad T^{-1}(\hmu)$. Recall that a risk-neutral agent moves the share vector so that the prices match its mean parameter. Therefore, define the \emph{target shares} as $\ttheta = \grad T^{-1}_{\lambda}(\hmu)$. The target shares $\htheta$ and natural parameter $\ttheta$ are related by
%
$$
\grad T(\htheta) = \grad T_{\lambda}(\ttheta) = \hmu.
$$
%
It is straightforward to check that $\grad T_{\lambda}(\ttheta) = \grad T(\lambda \ttheta)$ so we have
%
\begin{equation} \label{target}
\ttheta = \htheta / \lambda.
\end{equation}
%
Higher price responsiveness means fewer shares must be bought to make the market prices match the agent's expectation, so the natural parameter is scaled down accordingly. With a liquidity adjustment, the analysis of Theorem~\ref{thm:exp-util} can be extended and yields the following result.
%
\begin{corollary}
Under the conditions of Theorem~\ref{thm:exp-util} and an inverse liquidity of $\lambda$, the agent's optimal trade $\delta$ moves the state vector to
%
\begin{equation} \label{liqu-update}
\theta + \delta = \frac{\lambda}{\lambda+a} \ttheta + \frac{a}{\lambda+a} \theta
= \frac{\lambda}{\lambda+a} \grad T^{-1}_{\lambda}(\hmu) + \frac{a}{\lambda+a} \grad T^{-1}_{\lambda}(\mu).
\end{equation}
%
\end{corollary}
%
According to~(\ref{liqu-update}), as $\lambda$ grows large the agent moves the market state closer to the \emph{target shares}, rather than its true natural parameter. Note that the target shares themselves depend on $\lambda$ by~(\ref{target}), but the update can be directly written in terms of the agent's beliefs as in the right-hand side of~(\ref{liqu-update}).

@@

We will show that if we are able to impose finite initial budgets on the traders and control the market prices based on these budgets, then  it possible to set up the market so that it is prohibitive for damaging traders to participate in the market. Further, the informative traders can be shown to have expected growth in budget so that they are eventually able to move the market prices without restriction. This idea of a dual analysis has been used in a recommender system context to compute reputations in \cite{ressami07}.
